dispRity manual

### 2020-04-24 inst/gitbook/03_specific-tutorials.Rmd
book update (up until section 4)

The following section contains information specific to some functions.
If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!

@@

Here is an example for `method = discrete`:

@@

@@

@@

\begin{equation}
    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}
\end{equation}

\begin{equation}
    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
\end{equation}

\begin{equation}
    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
\end{equation}

@@

\begin{equation}
    displacements = \frac{\sqrt{\sum_{i=1}^{n}{({k}_{n}-Reference_{k})^2}}}{\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}}
\end{equation}

\begin{equation}
    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
\end{equation}

\begin{equation}
    n.ball.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R
\end{equation}

\begin{equation}
    radius = |\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|
\end{equation}

\begin{equation}
    ranges = |max(k_i) - min(k_i)|
\end{equation}

\begin{equation}
    variances = \sigma^{2}{k_i}
\end{equation}

@@

Where *k* is the number of dimensions,
*n* the number of elements,
$\Gamma$ is the Gamma distribution,
$\lambda_i$ is the eigenvalue of each dimensions,
$\sigma^{2}$ is their variance and
$Centroid_{k}$ is their mean,
$Ancestor_{n}$ is the coordinates of the ancestor of element $n$,
$f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.),
*R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid),
and $Reference_{k}$ is an arbitrary point's coordinates (usually 0).

@@

#### Orientation: angles

The `angle` performs a least square regression (via the `lm` function) and returns slope of the main axis of variation for each dimension. This slope can be converted into different units, `"slope"`, `"degree"` (the default) and `"radian"`. This can be changed through the `unit` argument. 
By default, the angle is measured from the slope 0 (the horizontal line in a 2D plot) but this can be changed through the `base` argument (using the defined `unit`):

@@

<!-- TODO: add significance -->

@@

The plots can be of four different types:

 * `continuous` for displaying continuous disparity curves
 * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.

@@

@@

## Note that no disarity has been calculated here:
is.null(cust_subsets$disparity)
is.null(time_subsets$disparity)

@@

@@

>>>>

The following section contains information specific to some functions.
If any of your questions are not covered in these sections, please refer to the function help files in `R`, send me an email (<guillert@tcd.ie>), or raise an issue on [GitHub](https://github.com/TGuillerme/dispRity/issues).
The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!

@@

### Time-binning

Here is an example for the time binning method (`method = discrete`):

@@

### Time-slicing

@@

More details about the differences between these methods can be found in @time-slice.

@@

\begin{equation}
    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({d}_{n}-Ancestor_{n})^2}}
\end{equation}

\begin{equation}
    centroids = \sqrt{\sum_{i=1}^{n}{({d}_{n}-Centroid_{d})^2}}
\end{equation}

\begin{equation}
    diagonal = \sqrt{\sum_{i=1}^{d}|max(d_i) - min(k_i)|}
\end{equation}

@@

\begin{equation}
    displacements = \frac{\sqrt{\sum_{i=1}^{n}{({d}_{n}-Reference_{d})^2}}}{\sqrt{\sum_{i=1}^{n}{({d}_{n}-Centroid_{k})^2}}}
\end{equation}

\begin{equation}
    ellipse.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} (\lambda_{i}^{0.5})
\end{equation}

\begin{equation}
    n.ball.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} R
\end{equation}

\begin{equation}
    radius = |\frac{\sum_{i=1}^{n}d_i}{n} - f(\mathbf{v}d)|
\end{equation}

\begin{equation}
    ranges = |max(d_i) - min(d_i)|
\end{equation}

\begin{equation}
    variances = \sigma^{2}{d_i}
\end{equation}

@@

Where *d* is the number of dimensions,
*n* the number of elements,
$\Gamma$ is the Gamma distribution,
$\lambda_i$ is the eigenvalue of each dimensions,
$\sigma^{2}$ is their variance and
$Centroid_{k}$ is their mean,
$Ancestor_{n}$ is the coordinates of the ancestor of element $n$,
$f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.),
*R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid),
and $Reference_{k}$ is an arbitrary point's coordinates (usually 0).

@@

#### Orientation: angles and deviations

The `angles` performs a least square regression (via the `lm` function) and returns slope of the main axis of variation for each dimension. This slope can be converted into different units, `"slope"`, `"degree"` (the default) and `"radian"`. This can be changed through the `unit` argument. 
By default, the angle is measured from the slope 0 (the horizontal line in a 2D plot) but this can be changed through the `base` argument (using the defined `unit`):

@@

The `deviations` function is based on a similar algorithm as above but measures the deviation from the main axis (or hyperplane) of variation.
In other words, it finds the least square line (for a 2D dataset), plane (for a 3D dataset) or hyperplane (for a >3D dataset) and measures the shortest distances between every points and the line/plane/hyperplane.
By default, the hyperplane is fitted using the least square algorithm from `stats::glm`:

```{r}
## The distribution of the deviation of each point from the least square hyperplane
summary(dispRity(dummy_space), metric = deviations))
```

It is also possible to specify the hyperplane equation through the `hyperplane` equation. The equation must contain the intercept first and then all the slopes and is interpreted as $intercept + Ax + By + ... + Nd = 0$. For example, a 2 line defined as beta + intercept (e.g. $y = 2x + 1$) should be defined as `hyperplane = c(1, 2, 1)` ($2x - y + 1 = 0$).

```{r}
## The distribution of the deviation of each point from a slope (with only the two first dimensions)
summary(dispRity(dummy_space[, c(1:2)]), metric = deviations, hyperplane = c(1, 2, -1)))
```

Since both the functions `angles` and `deviations` effectively run a `lm` or `glm` to estimate slopes or hyperplanes, it is possible to use the option `significant = TRUE` to only consider slopes or intercepts that have a slope significantly different than zero using an `aov` with a significant threshold of $p = 0.05$.
Note that depending on your dataset, using and `aov` could be completely inappropriate!
In doubt, it's probably better to enter your `base` (for `angles`) or your `hyperplane` (for `deviations`) manually so you're sure you know what the function is measuring.

@@

The plots can be of five different types:

 * `preview` for a 2d preview of the trait-space. 
 * `continuous` for displaying continuous disparity curves
 * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.

@@

### `type = preview`

@@

## Note that no disparity has been calculated here:
is.null(cust_subsets$disparity)
is.null(time_subsets$disparity)

@@

For `dispRity` objects that do contain disparity data, the default option is to plot your disparity data.
However you can always force the `preview` option using the following:

```{r}
par(mfrow = c(2,1))
## Default plotting
plot(disparity_time_slices, main = "Disparity through time")
## Plotting with more arguments
plot(disparity_time_slices, type = "preview", main = "Two first dimensions of the trait space")
```

@@

## Disparity from multiple matrices (and multiple trees!) {#multi.input}

Since the version `1.4` of this package, it is possible to use multiple trees and multiple matrices in `dispRity` objects.
To use multiple matrices, this is rather easy: just supply a list of matrices to any of the `dispRity` functions and, as long as they have the same size and the same rownames they will be handled as a distribution of matrices.

```{r}
set.seed(1)
## Creating 3 matrices with 4 dimensions and 10 elements each (called t1, t2, t3, etc...)
matrix_list <- replicate(3, matrix(rnorm(40), 10, 4, dimnames = list(paste0("t", 1:10))),
                         simplify = FALSE)
class(matrix_list) # This is a list of matrices

## Measuring some disparity metric on one of the matrices
summary(dispRity(matrix_list[[1]], metric = c(sum, variances)))

## Measuring the same disparity metric on the three matrices
summary(dispRity(matrix_list, metric = c(sum, variances)))
```

As you can see, when measuring the sum of variances on multiple matrices, we now have a distribution of sum of variances rather than a single observed value.

Similarly as running disparity analysis using multiple matrices, you can run the `chrono.subsets` function using multiple trees.
This can be useful if you want to use a tree posterior distribution rather than a single consensus tree.
These trees can be passed to `chrono.subsets` as a `"multiPhylo"` object (with the same node and tip labels in each tree).
First let's define a function to generate multiple trees with the same labels and root ages:

```{r}
set.seed(1)
## Matches the trees and the matrices
## A bunch of trees
make.tree <- function(n, fun = rtree) {
    ## Make the tree
    tree <- fun(n)
    tree <- chronos(tree, quiet = TRUE,
                    calibration = makeChronosCalib(tree, age.min = 10, age.max = 10))
    class(tree) <- "phylo"
    ## Add the node labels
    tree$node.label <- paste0("n", 1:Nnode(tree))
    ## Add the root time
    tree$root.time <- max(tree.age(tree)$ages)
    return(tree)
}
trees <- replicate(3, make.tree(10), simplify = FALSE)
class(trees) <- "multiPhylo"
trees
```

We can now simulate some ancestral states for the matrices in the example above to have multiple matrices associated with the multiple trees.

```{r}
## A function for running the ancestral states estimations
do.ace <- function(tree, matrix) {
    ## Run one ace
    fun.ace <- function(character, tree) {
        results <- ace(character, phy = tree)$ace
        names(results) <- paste0("n", 1:Nnode(tree))
        return(results)
    }
    ## Run all ace
    return(rbind(matrix, apply(matrix, 2, fun.ace, tree = tree)))
}

## All matrices
matrices <- mapply(do.ace, trees, matrix_list, SIMPLIFY = FALSE)
```

Let's first see an example of time-slicing with one matrix and multiple trees.
This assumes that your tip values (observed) and node values (estimated) are fixed with no error on them.
It also assumes that the nodes in the matrix always corresponds to the node in the trees (in other words, the tree topologies are fixed):

```{r}
## Making three "proximity" time slices across one tree
one_tree <- chrono.subsets(matrices[[1]], trees[[1]],
                           method = "continuous", model = "proximity", time = 3)
## Making three "proximity" time slices across the three trees
three_tree <- chrono.subsets(matrices[[1]], trees,
                             method = "continuous", model = "proximity", time = 3)
## Measuring disparity as the sum of variances and summarising it
summary(dispRity(one_tree, metric = c(sum, variances)))
summary(dispRity(three_tree, metric = c(sum, variances)))
```

This results show the effect of considering a tree distribution: in the first case (`one_tree`) the time slice at 3.95 Mya has a sum of variances of 2.9 but this values goes down to 0.256 in the second case (`three_tree`) which is due to the differences in branch lengths distributions:

```{r, print = FALSE}
par(mfrow = c(3,1))
slices <- c(7.9, 3.95, 0)
fun.plot <- function(tree) {
  plot(tree)
  nodelabels(tree$node.label, cex = 0.8)
  axisPhylo()
  abline(v = tree$root.time - slices)
}
silent <- lapply(trees, fun.plot)
```

Note that in this example, the nodes are actually even different in each tree! The node `n4` for example, is not direct descendent of `t4` and `t6` in all trees!
To fix that, it is possible to input a list of trees and a list of matrices that correspond to each tree in `chrono.subsets` by using the `bind.data = TRUE` option.
In this case, the matrices need to all have the same row names and the trees all need the same labels as before:

```{r}
## Making three "proximity" time slices across three trees and three bound matrices
bound_data <- chrono.subsets(matrices, trees,
                             method = "continuous", model = "proximity",
                             time = 3, bind.data = TRUE)
## Making three "proximity" time slices across three trees and three matrices
unbound_data <- chrono.subsets(matrices, trees,
                               method = "continuous", model = "proximity",
                               time = 3, bind.data = FALSE)

## Measuring disparity as the sum of variances and summarising it
summary(dispRity(bound_data, metric = c(sum, variances)))
summary(dispRity(unbound_data, metric = c(sum, variances)))
```

Note here that the results are again rather different: with the bound data, the slices are done across the three trees and each of their corresponding matrix (resulting in three observation) which is more accurate than the previous results from `three_trees` above.
With the unbound data, the slices are done across the three trees and applied to the three matrices (resulting in 9 observations).
As we've seen before, this is incorrect in this case since the trees don't have the same topology (so the nodes selected by a slice through the second tree are not equivalent to the nodes in the first matrix) but it can be useful if the topology is fixed to integrate both uncertainty in branch length (slicing through different trees) and uncertainty from, say, ancestral states estimations (applying the slices on different matrices).

### 2020-04-24 inst/gitbook/index.Rmd
book update (up until section 4)

## To CRAN or not to CRAN? {#noCRAN}

This package can be installed via [the CRAN](https://cran.r-project.org/package=dispRity) or via [GitHub](https://github.com/TGuillerme/dispRity).

 * The CRAN version is a more "strict" version of the package that does not include some functionalities such as this gitbook vignette because of compatibility issues with the CRAN requirement.
 * The GitHub version is a more "reactive" version (fed by user's suggestions and comments) and can contain more functionalities (see list below). This version exists on a `"release"` branch which is tested as strictly as the CRAN version with Travis and codecov. The `"master"` branch contains always the latest "piping hot", not yet release version. It might (though rarely) be bugged.

Note however, that apart from the functions listed below, both versions of the package have the same tested and implemented functions.

### So which version do I choose? {#version}

@@

<!-- ## Associated papers -->

>>>>

## Which version do I choose? {#version}

@@

### Why is it important to cite us?

Aside from how science works (if you're using a method from a specific, cite that specific paper to refer to that specific method), why is it important to also cite the package and the manual?

All the people involve in making the `dispRity` package happen do it enthusiastically and freely and most amazingly not asking anything in return!
I created the package with this idea in mind and I am still sticking to it.
However, academia (the institutions and people producing science around the globe) is unfortunately not optimal at many level (some might even say "broken"): high impact papers attract big grants that attract high impact papers and big grants again, all this along with livelihood, permanent position and job security.
Unfortunately however, method development has a hard time to catch up with the current publish or perish system: constantly updating the `dispRity` package and this manual is hugely time consuming (but really fun!) and that is not even taking into account maintenance and helping users.
Although I do truly believe that this time spent doing these things modestly help the scientific endeavour, it does not contribute to our paper list!

Therefore, by citing the package and this manual, you help provide visibility to other workers and you might help them in their work!
And you directly contribute in making this project fun for all the people involved and most of all, free, updated and independent from the publish and perish system!

Thank you!
<!-- ## Associated papers -->
